---
title: "Reading and translating genomic data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading and translating genomic data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
library(genometranslator)
```

## Overview

`genometranslator` separates data translation from data filtering. Reader
functions import genomic files and standardize their sample, marker, and
genotype information. Writer functions translate those data to another format;
they do not silently filter markers or individuals. The user is responsible for
quality control and for preparing data that are appropriate for the intended
downstream analysis.

There are two ways to read or write data:

* `read_genome()` and `write_genome()` provide a simple interface and select a
  format-specific function automatically;
* functions such as `read_vcf()`, `read_dart()`, and `write_bayescan()` expose
  arguments specific to a particular format.

The package uses a GDS-first workflow. Readers return a file-backed GDS for
formats such as VCF, DArT, and PLINK, allowing large datasets to be inspected,
filtered, and translated without first creating one row per sample and marker
in memory. Use `tidy_genome()` only when a downstream task specifically needs
an in-memory table.

The former `tidy_vcf()` convenience wrapper is no longer part of
`genometranslator`. Use `read_genome()` for automatic format detection or
`read_vcf()` for VCF-specific import controls. Apply genomic filters with
`radr`, and call `tidy_genome()` separately if a table is required. The legacy
`tidy_vcf()` implementation remains in `radiator` for historical workflows.

The examples below are not evaluated because they refer to files supplied by
the user.

**It all starts with a well-maintained sample metadata file.**

## Maintain the sample metadata first

The sample metadata table is the authoritative record describing the samples
used in a project. It can be maintained in Excel, LibreOffice, Google Sheets,
R, a database, or another suitable tool. The software matters less than using a
consistent tabular structure and preserving the file carefully.

The tidy-data principles emphasized throughout
[R for Data Science](https://r4ds.hadley.nz) by Hadley Wickham,
Mine Çetinkaya-Rundel, and Garrett Grolemund are particularly useful:

* each row represents one sample;
* each column represents one variable;
* each cell contains one value;
* each type of observational unit is stored in its own table.

For example, sampling site, collection year, sex, sequencing batch, original
sample identifier, standardized sample identifier, and DArT `TARGET_ID` should
be separate columns. Do not combine several variables in one cell, use merged
cells or multiple header rows, or encode information only with colours, fonts,
or comments. Column names should be unique and stable, and missing values should
be represented consistently.

When using spreadsheet software, format identifier columns as text so that
leading zeros are preserved. Avoid formulas in the final exchange table and
export a tab-delimited text file for use with R. Keep the complete spreadsheet
or metadata table as the long-term record; exported strata files can always be
regenerated from it.

The metadata table is also the authoritative place to document original and
standardized identifiers, along with the reason or date for a name change. The
strata file should not be used to maintain that history.

## Prepare the strata file

Once the metadata are in good order, generate the strata file needed for the
genomic import or analysis. This small step prevents many difficult-to-diagnose
matching problems later. A strata file is a compact, tab-delimited table with
one row per sample and, for most formats, two required columns:

* `INDIVIDUALS` is the unique, stable name used for each sample;
* `STRATA` is the group to which the sample belongs for the current analysis.

For DArT data, a third column named `TARGET_ID` connects the names found in the
original DArT report to the standardized names in `INDIVIDUALS`.

### Standardize `INDIVIDUALS`

Good sample identifiers are meaningful, unique, and stable. Avoid identifiers
such as `1`, `2`, and `3`: they contain no context, are easily confused with row
numbers, and may change when a table is sorted. Prefer an identifier such as
`GWH-TAS-2025-001`, where each component has a documented meaning.

Useful practices include:

* record the identifier found in the genotype file and its standardized
  replacement in the sample metadata table;
* join existing metadata columns when necessary, for example species, sampling
  site, year, and a within-site sample number;
* use the same component order and separator for every sample;
* prefer hyphens (`-`) as separators; underscores can become visually unclear
  when identifiers are underlined or printed, and periods (`.`) can be altered
  by some R functions;
* avoid spaces, accented characters, slashes, and other punctuation;
* keep capitalization consistent and preserve meaningful leading zeros;
* verify that identifiers are unique, non-missing, and match the genotype data
  exactly.

Suppose the complete sample metadata are stored in `sample_metadata.tsv` with
columns named `SPECIES`, `SITE`, `YEAR`, `SAMPLE_NUMBER`, and `REGION`. A compact
strata file can be generated as follows:

```{r create-strata, eval=FALSE}
sample_metadata <- readr::read_tsv(
  "sample_metadata.tsv",
  col_types = readr::cols(.default = readr::col_character())
)

strata <- sample_metadata |>
  dplyr::mutate(
    INDIVIDUALS = paste(SPECIES, SITE, YEAR, REGION, SAMPLE_NUMBER, sep = "-"),
    STRATA = REGION
  ) |>
  dplyr::select(INDIVIDUALS, STRATA)

stopifnot(
  !anyNA(strata$INDIVIDUALS),
  !anyDuplicated(strata$INDIVIDUALS),
  !anyNA(strata$STRATA)
)

readr::write_tsv(strata, "strata.tsv")
```

Read identifier-like columns as character data so that values such as `001` do
not become `1`. Inspect the result before importing the genomic data:

```{r inspect-strata, eval=FALSE}
strata
dplyr::count(strata, STRATA)
```

### Renaming samples while reading a VCF

For a VCF, sample identifiers can be changed during import by adding `NEW_ID`
to the strata file. In this case, `INDIVIDUALS` must match the current sample
identifier in the VCF exactly, and `NEW_ID` supplies the standardized identifier
that will be written to the GDS object:

```text
INDIVIDUALS  NEW_ID           STRATA
Shark_1      GWH-TAS-2025-001 TAS
Shark_2      GWH-VIC-2025-001 VIC
Shark_3      GWH-TAS-2025-002 TAS
```

Generate this file from the authoritative metadata table rather than entering
the correspondence manually in the strata file:

```{r create-vcf-strata, eval=FALSE}
vcf_strata <- sample_metadata |>
  dplyr::transmute(
    INDIVIDUALS = VCF_ID,
    NEW_ID = STANDARD_ID,
    STRATA = REGION
  )

stopifnot(
  !anyNA(vcf_strata),
  !anyDuplicated(vcf_strata$INDIVIDUALS),
  !anyDuplicated(vcf_strata$NEW_ID)
)

readr::write_tsv(vcf_strata, "vcf_strata.tsv")
```

Here, `VCF_ID` and `STANDARD_ID` are columns maintained in
`sample_metadata.tsv`. If the identifiers already used in the VCF are suitable,
omit `NEW_ID` and use the regular two-column strata file.

### Keep `STRATA` simple

`STRATA` represents the grouping needed for the current analysis, not the
entire sample metadata record. Short acronyms such as `NTH` and `STH`, `POP1`
and `POP2`, or site codes such as `TAS` and `VIC` are preferable. Long labels
can make axes, legends, and facet labels unreadable when figure space is tight.
Use a single consistent vocabulary: do not mix values such as `STH`, `South`,
and `south`.

If a hierarchical label is needed, use a documented and consistent structure
such as `TAS-HOB` and `TAS-FRE`. Avoid packing every available variable into
`STRATA`. Keep collection date, sex, life stage, coordinates, sequencing batch,
and other descriptive fields in the complete sample metadata table. They can be
joined to analysis results later using `INDIVIDUALS` as the key.

Also check the following before saving the file:

* every row represents one sample;
* every sample needed for the analysis occurs exactly once;
* samples absent from the strata file are intentionally excluded;
* group names do not contain trailing spaces or inconsistent separators;
* the file is written as tab-delimited text rather than an Excel workbook.

A plain-text editor that can display invisible characters is useful for the
final inspection. For example, BBEdit can reveal tabs, spaces, and line endings,
making accidental spaces, mixed delimiters, and hidden characters easier to
find.

## Reading a VCF file

Suppose that `individuals.vcf.gz` contains the genotypes and `strata.tsv`
contains sample assignments. A minimal strata table has one row per individual:

```text
INDIVIDUALS  STRATA
GWH-TAS-2025-001  TAS
GWH-VIC-2025-001  VIC
GWH-TAS-2025-002  TAS
```

Use `read_vcf()` when VCF-specific control is wanted:

```{r read-vcf, eval=FALSE}
shark <- read_vcf(
  data = "individuals.vcf.gz",
  strata = "strata.tsv",
  filename = "shark",
  parallel.core = 8,
  verbose = TRUE
)
```

The function returns an open SeqArray GDS object. The GDS representation avoids
materializing a very large genotype table in memory and is the preferred format
for subsequent work in R.

### Import variant and invariant sites

Use `read_vcf("populations.all.vcf", all.sites = TRUE)` for a site-level
all-sites VCF. The default `all.sites = NULL` detects invariant records;
`FALSE` rejects them. Import preserves missing genotypes rather than treating
them as reference calls. The GDS retains existing caller/version provenance,
adds `vcf.site.content` with import-time record counts, and adds `SITE_TYPE`
to marker metadata. An ALT-bearing record can still be monomorphic among
selected individuals, so this classification is not an allele-frequency test.

STACKS `--vcf-all`, FreeBayes `--report-monomorphic`, and bcftools `call -m`
without `-v` can supply such records. In scaffoldr, set `all.sites = TRUE`
in `run_freebayes()` or `run_bcftools_mpileup()`. This requests site-level
output, not gVCF reference blocks. Caller thresholds and missingness still
limit what is observed; an all-sites tag does not guarantee genome-wide
callability. STACKS invariant records may contain GT only, without DP or GQ.

Keep an SFS-specific dataset: sample and locus quality checks remain useful,
but removing invariant records or applying MAF/MAC thresholds changes the SFS.
Both `write_fastsimcoal()` and `write_dadi()` count eligible invariant records
directly from the GDS. They require complete calls across selected samples;
the all-sites tag alone does not certify callability. Blocked gVCF remains
future work.

### Preserve how variants were generated

`read_vcf()` stores a provenance table in the GDS metadata. It records the
original path, size, modification time, MD5 checksum, import time, detected
caller, exact VCF source value, imported INFO and FORMAT fields, and available
reference, contig, version, and command header lines. Keep the original VCF and
its workflow files as well: provenance stored in GDS supports interpretation,
but it is not a replacement for a maintained variant-calling pipeline.

This matters when a genome-scan peak, missingness pattern, or local PCA signal
is restricted to one caller, mapping strategy, sequencing batch, or reference
assembly. Such a pattern can be biological, technical, or both.

### Structural variants are not ordinary SNP dosages

When present, common annotations such as `SVTYPE`, `END`, `SVLEN`, `CIPOS`,
`CIEND`, `MATEID`, and `EVENT` are preserved during VCF import. This is the
first step toward fuller structural-variant support. Structural variants can
have uncertain breakpoints, symbolic ALT alleles, breakends, copy number, and
several forms of read support, so they should not all be forced into the SNP
`ALT_DOSAGE` representation.

Until a dedicated structural-variant representation is complete, keep these
annotations and the source VCF, and describe local signals cautiously. A local
PCA or LD haploblock may be consistent with a candidate inversion, but it may
also reflect a centromere, low recombination, assembly or mapping problems,
introgression, population-specific missingness, or another structural variant.
Use "candidate", "putative", or "inversion-associated haploblock" until
physical evidence supports a more specific structural interpretation.

VCF files that are not ready for parallel access may need to be sorted,
bgzip-compressed, and indexed. These preparation steps can require the optional
`bcftools` executable. Check its visibility with:

```{r check-bcftools, eval=FALSE}
Sys.which("bcftools")
genometranslator_dependencies()
```

Use `tidy_genome()` only when an in-memory tidy table is needed:

```{r vcf-to-tidy, eval=FALSE}
shark_tidy <- tidy_genome(
  data = shark,
  parallel.core = 8
)
```

Do not convert a whole-genome GDS merely as an import step. A tidy table has
approximately one row per sample-marker combination and can therefore be much
larger in memory than the file-backed GDS. Extract or materialize only the data
needed by the downstream analysis.

## Reading DArT files

DArT reports have changed over time, and the package recognizes several related
layouts automatically:

* **1row**: one row per SNP marker, with genotypes encoded as `0`, `1`, `2`, or
  missing;
* **2rows**: two rows per marker representing presence or absence of the REF and
  ALT alleles;
* **counts**: two rows per marker containing read counts or depth for the REF and
  ALT alleles rather than genotype calls;
* **silico.dart**: dominant SilicoDArT presence/absence markers without REF and
  ALT alleles;
* **silico.dart.counts**: SilicoDArT presence/absence information accompanied by
  sequence counts.

For DArT input, prepare a three-column strata file. `TARGET_ID` must match the
identifier in the original DArT report, while `INDIVIDUALS` contains the stable,
standardized name that will be used after import:

```text
TARGET_ID  INDIVIDUALS       STRATA
1002451    GWH-TAS-2025-001  TAS
1002452    GWH-TAS-2025-002  TAS
1002453    GWH-VIC-2025-001  VIC
```

When the DArT identifiers are difficult to locate, extract them directly from
the report instead of editing the large genotype file manually:

```{r dart-target-id, eval=FALSE}
dart_ids <- extract_dart_target_id(
  data = "dart_counts.csv",
  write = FALSE
)
```

For DArT files, `INDIVIDUALS` must be generated by the user. This is the right
time to create meaningful, standardized identifiers. After verifying which
sample corresponds to each `TARGET_ID`, store that relationship permanently in
the complete sample metadata table. Then use the `TARGET_ID` values extracted
from the DArT report to select and join the corresponding metadata records. The
DArT strata file is a small export of the three fields needed for import:

```{r create-dart-strata, eval=FALSE}
unmatched_dart_ids <- dart_ids |>
  dplyr::anti_join(sample_metadata, by = "TARGET_ID")

stopifnot(nrow(unmatched_dart_ids) == 0L)

dart_strata <- dart_ids |>
  dplyr::left_join(sample_metadata, by = "TARGET_ID") |>
  dplyr::transmute(
    TARGET_ID,
    INDIVIDUALS = STANDARD_ID,
    STRATA = REGION
  )

stopifnot(
  !anyNA(dart_strata),
  !anyDuplicated(dart_strata$TARGET_ID),
  !anyDuplicated(dart_strata$INDIVIDUALS)
)

readr::write_tsv(dart_strata, "dart_strata.tsv")
```

Never infer this correspondence from row or column order alone. The optional
metadata inspection features of `extract_dart_target_id()` are described in
that function's documentation.

`detect_genomic_format()` identifies the file as DArT, while
`detect_dart_format()` reports the particular DArT layout:

```{r detect-dart, eval=FALSE}
detect_genomic_format("dart_counts.csv")
detect_dart_format("dart_counts.csv")
```

For a DArT count file, use:

```{r read-dart-counts, eval=FALSE}
dart_genome <- read_dart(
  data = "dart_counts.csv",
  strata = "dart_strata.tsv"
)
```

The specialized `read_dart()` documentation describes additional DArT columns,
metadata, and calibration options. Calling `read_dart()` directly is
recommended when those choices need to be controlled.

### Stable marker identifiers and coordinates

Imported marker metadata deliberately separates identity from genomic placement.
`MARKERS` is a stable identifier with three components separated by two
underscores. A de novo marker is written as
`DENOVO__<LOCUS>__<POS>`. For a reference marker, the exact sequence name is
kept in `CHROM`, while only the component embedded in `MARKERS` is made safe for
filenames and downstream code. For example, `CHROM = "NC_054479.1"` is encoded
as `NC_054479_1` inside `MARKERS`.

Consequently, genomic coordinates must always be taken from the `CHROM` and
`POS` columns, not reconstructed by splitting `MARKERS`. `LOCUS` retains the
original DArT/read locus and `COL` retains the SNP position within that source
sequence. Sanitizing two distinct chromosome names to the same marker component
is treated as an error rather than silently creating ambiguous identifiers.

## The generic read and write interface

When format-specific options are not required, `read_genome()` detects the
input and calls the corresponding reader with its defaults. Only the shared
`strata` argument needs to be supplied:

```{r read-genome, eval=FALSE}
genome <- read_genome(
  data = "individuals.vcf.gz",
  strata = "strata.tsv",
  parallel.core = 8,
  verbose = TRUE
)
```

The same entry point can be tried with a DArT, FSTAT, Genepop, GDS, PLINK, or
package-native genomic file, and with `genind`, `genlight`, or `gtypes` objects:

```{r read-genome-dart, eval=FALSE}
dart_genome <- read_genome(
  data = "dart_counts.csv",
  strata = "dart_strata.tsv"
)
```

`write_genome()` dispatches to a format-specific writer. Specify `output`
explicitly when the filename extension is ambiguous:

```{r write-genome, eval=FALSE}
write_genome(
  data = genome,
  output = "genepop",
  filename = "shark",
  verbose = TRUE
)

write_genome(
  data = genome,
  output = "vcf",
  filename = "shark_filtered",
  verbose = TRUE
)
```

The name `shark_filtered.vcf` should only be used if `genome` was filtered
before the call: the writer itself does not perform that filtering.

### Export VCF or BCF

The filename selects the encoding; no separate BCF writer is needed:

```{r vcf-bcf-export, eval=FALSE}
write_vcf(genome, filename = "study.vcf")
write_vcf(genome, filename = "study.vcf.gz")
write_vcf(genome, filename = "study.bcf")
```

Compressed VCF and BCF require bcftools, are coordinate-sorted, and receive
a CSI index by default (`index = FALSE` disables indexing). Plain VCF keeps
the GDS record order. Existing output requires `overwrite = TRUE`.
GDS selections and metadata whitelists are respected and restored, and
available genotype fields (including missing calls, depth and likelihoods)
are retained. Missing annotations cannot be recovered by changing encoding.
Original site-level INFO summaries are not recalculated when samples change;
recalculate cohort-dependent AF/AC/AN/NS/DP before interpreting them.
BCF is binary VCF, not blocked gVCF. Legacy tidy input remains supported,
but only through its existing limited genotype-table translation.

### Export an Arlequin project

`write_arlequin()` reads active diploid genotypes directly from GDS and writes
an `.arp` project. It supports biallelic and multiallelic records, writes missing
alleles as `?`, and assigns frequency one to each individual. Each VCF record
becomes one unphased `STANDARD` locus. Original sample/population names and
allele definitions remain available in companion mapping tables.

```{r arlequin-export, eval=FALSE}
arp <- write_arlequin(
  genome, filename = "study", path.folder = "exports",
  strata = strata, pop.levels = c("North", "South")
)
arp$files
```

Population order must list every exported population; it never silently
removes samples. The writer respects active selections and metadata
whitelists, restores temporary selections even on errors, and leaves the
caller's GDS connection open. The project places populations in one structural
group; this is not an inferred hierarchy. See Excoffier, Laval, and Schneider
(2005), *Arlequin ver. 3.0: An integrated software package for population
genetics data analysis*, **Evolutionary Bioinformatics Online**, 1, 47-50,
and the [official format specification](https://cmpg.unibe.ch/software/arlequin3522/man/Arlequin35.pdf).

### Export a site-frequency spectrum for fastsimcoal

`write_fastsimcoal()` builds a joint spectrum from active diploid biallelic
SNPs. The default folded spectrum does not require ancestral states. Population
order must match the demographic model; the population table records zero-based
indices and chromosome sample sizes. The last population varies fastest in the
flattened file, following the
[fastsimcoal specification](https://cmpg.unibe.ch/software/fastsimcoal2/man/fastsimcoal28.pdf).

```{r fastsimcoal-export, eval=FALSE}
sfs <- write_fastsimcoal(
  genome, strata = strata, pop.levels = c("North", "South"),
  filename = "study", path.folder = "exports", folded = TRUE
)
sfs$populations
table(sfs$audit$STATUS)
```

Use the resulting `_MSFS.obs` file with `--multiSFS` and `-m`; an unfolded
`_DSFS.obs` uses `--multiSFS` and `-d`. For unfolded output, supply a validated
`VARIANT_ID`/`ANCESTRAL` table. REF is not automatically ancestral.

Important limits and decisions:

- Only records with complete calls in every selected sample contribute.
  Incomplete, non-biallelic/non-SNP, and unknown-ancestry exclusions are audited.
  This initial exporter does not perform projection or genotype-likelihood SFS
  estimation. Nonrandom missingness can bias complete-case spectra.
- Folding identifies the minor allele globally across the selected populations.
  Pooled ties contribute half to each complementary cell.
- `invariant.sites` adds only independently validated, callable invariant sites
  **absent from the GDS**. Never count filtered SNPs, already represented sites,
  or uncallable sequence as additional invariant sites.
- Invariant records already in an all-sites GDS are counted automatically.
  For unfolded spectra, they also need ancestral bases. A REF base different
  from the supplied ancestral base contributes to the fixed-derived bin.
  The audit records both input site type and observed invariance after sample
  selection. Duplicate mapped positions are excluded to avoid double counting.
- A SNP-only file is not a complete sequence SFS. Account for variant discovery,
  MAC/MAF filters, RAD locus ascertainment, linkage, and callable sequence before
  demographic inference. Absolute scaling needs an appropriate denominator;
  no chromosome-length shortcut is used here.
- Joint spectra grow rapidly with sample size and number of populations. The
  `max.cells` limit prevents unexpectedly large allocations.

### Export a native dadi spectrum

`write_dadi()` now accepts GDS only and writes dadi's native `.fs` format,
not the former tidy-data SNP table. It shares the fastsimcoal SFS engine,
including invariant accounting, ancestry checks and complete-case exclusions.

```{r dadi-export, eval=FALSE}
dadi_sfs <- write_dadi(
  genome, strata = strata, pop.levels = c("North", "South"),
  filename = "study", path.folder = "exports", folded = TRUE
)
dadi_sfs$parameters
```

Read the result in Python with `dadi.Spectrum.from_file("exports/study.fs")`.
Population IDs P1, P2, etc. map to original names in the population sidecar.
The all-ancestral and all-derived corners are masked by default: counts are
preserved but excluded from ordinary dadi inference. Folded spectra also
mask cells above half the pooled chromosome count. For a model explicitly
using monomorphic bins, both export and import mask settings must be chosen
accordingly (`mask.corners = FALSE` in R; `mask_corners=False` in Python).
See the [dadi format specification](https://dadi.readthedocs.io/en/latest/user-guide/importing-data/).

Neither exporter imputes missing calls, projects sample sizes, or interprets
blocked gVCF. These remain explicit future work. With many individuals,
complete-case filtering can be severe: inspect the audit rather than assuming
the resulting SFS represents all sequenced loci.

### Upgrade legacy `.gds.rad` files

New GDS files use the standard `.gds` extension regardless of whether their
source was DArT, VCF, PLINK, or another format. The source format and provenance
belong in the GDS metadata rather than in the filename.

Use `upgrade_genome_gds()` to migrate an older package or plain SeqArray GDS:

```{r upgrade-gds, eval=FALSE}
new_file <- upgrade_genome_gds("shark.gds.rad")
# Writes shark.gds and shark_upgrade_report.tsv
```

The original file is never modified. The upgrade copies the file atomically,
migrates `/radiator` metadata to `/genometranslator`, creates current-schema
nodes, reconstructs marker and individual metadata where possible, modernizes
legacy marker identifiers, inventories depth and likelihood nodes, and checks
metadata identifiers against the core SeqArray variants and samples. Previous
marker identifiers are retained in `ORIGINAL_MARKERS` when they change.

Information that was never stored cannot be recreated. The report therefore
distinguishes migrated, reconstructed, retained, unavailable, and failed or
warning checks. In particular, missing strata, source sequences, DP, AD, or
provenance must be supplied from the original files if they are required.

For a dry metadata-preserving conversion without identifier modernization, use:

```{r upgrade-gds-conservative, eval=FALSE}
new_file <- upgrade_genome_gds(
  "shark.gds.rad",
  modernize.markers = FALSE
)
```

When finished with an open GDS connection, close it with:

```{r close-genome, eval=FALSE}
write_genome(genome)
```

With no output requested, `write_genome()` retains its GDS-management behavior:
it synchronizes and closes the open connection.

## Translating in one step

`genome_translator()` combines `read_genome()` and `write_genome()`. For
example, translate a VCF directly to Genepop using the default behavior of both
format-specific functions:

```{r genome-translator, eval=FALSE}
genome_translator(
  data = "individuals.vcf.gz",
  strata = "strata.tsv",
  output = "genepop",
  filename = "shark",
  parallel.core = 8,
  verbose = TRUE
)
```

This compact interface is appropriate when the defaults are suitable. For more
control, split the operation into a specialized reader and writer:

```{r specialized-translation, eval=FALSE}
genome <- read_vcf(
  data = "individuals.vcf.gz",
  strata = "strata.tsv",
  vcf.stats = TRUE,
  parallel.core = 8
)

write_genepop(
  data = genome,
  filename = "shark"
)

write_genome(genome) # synchronize and close the GDS connection
```

Always consult the selected reader and writer documentation for format-specific
dependencies, validation rules, and data-preparation recommendations.

### Genetic stock identification exports

`write_gsi_sim()` and `write_rubias()` accept diploid GDS files or open GDS
objects. They respect the active selections and metadata whitelists without
performing additional biological filtering. Both read genotypes in sample blocks,
support multiallelic loci, and preserve allele identities in mapping tables.
Neither format retains phase, coverage, or genotype likelihoods.

```{r gsi-export, eval=FALSE}
gsi <- genometranslator::write_gsi_sim(
  "study.gds", strata = "samples.tsv", filename = "baseline",
  path.folder = "exports"
)
r <- genometranslator::write_rubias(
  "study.gds", strata = "samples.tsv", filename = "study",
  path.folder = "exports"
)
```

For an all-reference rubias dataset, metadata needs `INDIVIDUALS` and `STRATA`.
For reference and mixture samples together, supply `INDIVIDUALS`,
`SAMPLE_TYPE`, `REPUNIT`, and `COLLECTION`. Reference collections must each
belong to one reporting unit. Mixtures require `REPUNIT = NA` and a distinct
collection name identifying their sampling group. Export them together, then
separate the returned rows: this keeps allele coding and locus columns aligned.
The genetic columns start at column 5. The returned rubias tibble must fit in
memory despite block-wise GDS reads.

The gsi_sim writer streams a text file with safe sample, population and locus
identifiers, plus mapping TSV files. Its return value is now a list; the genotype
file is `gsi$files[1]`. Export baseline and mixture inputs separately with the
same marker selection and allele dictionaries. Fully missing calls are zero pairs
in gsi_sim and NA pairs in rubias; partially missing diploid calls are rejected
by both writers. Existing files are protected unless `overwrite = TRUE`.
These writers now require GDS input; import other formats with `read_genome()`
first. The unused `parallel.core` argument was removed from `write_rubias()`.

See the [official gsi_sim documentation](https://github.com/eriqande/gsi_sim)
and [rubias input guide](https://eriqande.github.io/rubias/articles/rubias-overview.html).
The method references are Anderson, Waples and Kalinowski (2008),
*Canadian Journal of Fisheries and Aquatic Sciences* 65:1475–1486;
Anderson (2010), *Molecular Ecology Resources* 10:701–710; and
Moran and Anderson (2019), *Canadian Journal of Fisheries and Aquatic Sciences*
76:551–560, [doi:10.1139/cjfas-2018-0016](https://doi.org/10.1139/cjfas-2018-0016).
These tools prepare input, not an analysis: reference design, batch effects,
linked loci and marker ascertainment still need careful consideration.

### Genepop exchange

`read_genepop()` validates diploid Genepop files, including population separators,
individual records continued across lines, and two- or three-digit allele codes
(four or six digits per diploid genotype). Coding widths may differ between loci,
but must be consistent within a locus. Haploid records are explicitly rejected.
Missing allele copies are retained as `000`; they are not converted to observed
alleles. The result is a long tibble by default, or a wide table with `tidy = FALSE`.

`write_genepop()` reads the active GDS selection in sample blocks, without first
building a full long genotype table. It also accepts validated long tables with
`STRATA`, `INDIVIDUALS`, `MARKERS`, and `GT`. Population assignments can be supplied
with `strata` for GDS input. Files are protected against accidental overwriting.

```{r genepop-exchange, eval=FALSE}
export <- genometranslator::write_genepop(
  data = "study.gds", strata = "samples.tsv", filename = "study",
  path.folder = "exports", chunk.size = 32L
)
calls <- genometranslator::read_genepop(export$files[1])
```

The writer uses safe sample and locus names and writes companion sample,
population, locus, and allele mapping tables. GDS alleles receive numeric labels
(`001`, `002`, and so on); these labels are not microsatellite repeat lengths.
Genepop supports multiallelic loci, but this export does not retain phasing,
coverage, genotype likelihoods, or reference coordinates in the genotype file.
Keep the GDS and companion tables for those purposes. See the
[official Genepop input specification](https://f-rousset.r-universe.dev/genepop/doc/the-input-file.html)
for the format rules.

### Exporting to kinference

`write_kinference()` saves a `gbasics::snpgeno` object as RDS, together with
sample, locus and allele mapping tables. It preserves the active GDS selection
and matches optional sample metadata by `INDIVIDUALS`, retaining columns such
as sampling year, age, sex and plate. These covariates matter for interpretation
and subsequent CKMR models; providing them does not adjust the genetic analysis
for population structure or batch effects.

```{r kinference-export, eval=FALSE}
kin <- genometranslator::write_kinference(
  data = "study.gds", strata = "samples.tsv",
  filename = "study", path.folder = "exports"
)
# No-null model, after appropriate population and genotype QC:
kin <- kinference::est_ALF_nonulls(kin)
kin <- kinference::kin_power(kin, k = 0.5)
```

This first exporter supports complete, polymorphic, diploid biallelic nucleotide
SNPs. REF maps to A and ALT to B; they are not necessarily major/minor alleles.
The standard four-way labels AAO, AB and BBO are used, but no biological nulls
are inferred. Ordinary missing calls are **not** OO double-nulls. Missing or
partial calls therefore stop export; they are never automatically imputed,
discarded, or converted to biological absence. Depth and likelihoods remain in
the original GDS. The output and conversion matrices must fit in memory.

kinference is a specialised kin-finding and QC toolkit for Close-Kin
Mark-Recapture, not a demographic-model fitter or an automated pedigree builder.
Its strengths include attention to false-positive/false-negative rates, null
alleles, and the effect of linkage on close-kin statistics without requiring a
reference assembly. Its assumptions still matter: high-quality calls, a
genetically homogeneous population, approximately HWE loci, and limited LD.
Accounting empirically for physical linkage does not correct plate effects or
make dense, strongly LD-correlated whole-genome panels suitable without thinning.
Select the population and markers deliberately before export. Complete-case
filtering itself can alter ascertainment and should also be assessed.

In kinference 1.2.48, testing identified loss of character pair IDs in the
duplicate and HSP finders, and a resulting error in the POP finder.
Use `ij_numeric = TRUE` and map indices back through
`kin$info$INDIVIDUALS`. The upstream automatic-threshold implementation also
requires caution: its range check and variance-estimation subset can fail or
truncate the data used to estimate the HSP variance. Do not treat a returned
threshold as validated solely because the function completed.

Reference: Bravington MV, Baylis SM, Eveson JP, Feutry P (2026).
*kinference: Pairwise kinship detection for Close-Kin Mark-Recapture*.
bioRxiv preprint, [doi:10.64898/2026.05.18.725841](https://doi.org/10.64898/2026.05.18.725841).
See the [kinference manual and vignette](https://closekin.r-universe.dev/kinference)
and [gbasics input documentation](https://markbravington.r-universe.dev/gbasics).

### Other population-genetic exports and R objects

The following writers accept a SeqArray GDS path or open GDS object. They use
the active sample and marker selection and restore it even when export fails.
Import other formats first with `read_genome()`. Export does not silently remove
incompatible loci, impute missing calls, or infer phase. Allele-based writers
reject partially missing diploid calls; fully missing calls use the target
format's missing-value convention.

| Destination | Function | Important requirements |
|:--|:--|:--|
| FSTAT / hierfstat | `write_hierfstat()` | Diploid allele codes; supports multiallelic loci. |
| adegenet genind | `write_genind()` | Diploid codominant data; supports multiallelic loci. |
| adegenet genlight | `write_genlight()` | Biallelic ALT dosage. |
| strataG gtypes | `write_gtypes()` | Diploid allele pairs; optional strataG package. |
| STRUCTURE | `write_structure()` | Two rows per individual, locus header, missing `-9`. |
| fastSTRUCTURE | `write_faststructure()` | Biallelic; two rows and six metadata columns, or PLINK BED. |
| BayeScan | `write_bayescan()` | Population allele counts; rejects unobserved population–locus cells. |
| fineRADstructure / RADpainter | `write_fineradstructure()` | One variant record per locus with haplotype alleles; no unphased SNP concatenation. |
| HapMap | `write_hapmap()` | Biallelic nucleotide SNPs with real reference coordinates. |
| HZAR | `write_hzar()` | Biallelic frequencies and called sample counts; explicit population distances. |
| LDna | `write_ldna()` | Pairwise r² matrix; quadratic memory, with a marker-count guard. |
| pcadapt | `write_pcadapt()` | Locus-by-individual biallelic ALT dosage; missing `9`. |
| related | `write_related()` | Individual identifier and diploid allele pairs; missing `0`. |
| SNPRelate | `write_snprelate()` | Biallelic SNPRelate GDS; close the returned handle after use. |
| stockR | `write_stockr()` | Locus-by-individual biallelic dosage and sample groups. |

These are exchange formats, not substitutes for the original GDS. Depth,
likelihoods, phase, and annotations are not preserved by most targets. Safe
identifiers and allele encodings are documented in companion mapping tables;
object outputs also carry mapping attributes. Most conversions create an
in-memory matrix, so the selected dataset must fit in RAM. Blocked GDS reads
reduce temporary allocation but do not make the resulting R objects file-backed.

`read_fstat()`, `read_genind()`, `read_genlight()`, and `read_gtypes()` validate
their respective input representations. The object readers accept saved RDS
files. genind locus names containing periods are preserved, and genlight dosage
orientation is retained. For genind and genlight, requesting tidy output returns
the table without also creating a GDS; their GDS conversion currently requires
biallelic data. Multiallelic data can still be represented in the tidy result.

`radr::run_bayescan()` can export an active GDS selection, including population
subsamples. It validates numerical settings, quotes file paths, and stops on
external-process failure or missing results. Successful execution does not
establish that BayeScan's population model fits the study design.

`write_maverick()` has been removed. Use `write_structure()` when STRUCTURE input
is appropriate; software-specific settings still require independent checking.

Consult the [fastSTRUCTURE input documentation](https://github.com/rajanil/fastStructure),
[BayeScan documentation](https://github.com/mfoll/BayeScan),
[fineRADstructure documentation](https://www.milan-malinsky.org/fineradstructure),
and [pcadapt guide](https://bcm-uga.github.io/pcadapt/articles/pcadapt.html)
before running the downstream analysis.
