---
title: "Reading and translating genomic data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading and translating genomic data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
library(genometranslator)
```

## Overview

`genometranslator` separates data translation from data filtering. Reader
functions import genomic files and standardize their sample, marker, and
genotype information. Writer functions translate those data to another format;
they do not silently filter markers or individuals. The user is responsible for
quality control and for preparing data that are appropriate for the intended
downstream analysis.

There are two ways to read or write data:

* `read_genome()` and `write_genome()` provide a simple interface and select a
  format-specific function automatically;
* functions such as `read_vcf()`, `read_dart()`, and `write_bayescan()` expose
  arguments specific to a particular format.

The examples below are not evaluated because they refer to files supplied by
the user.

**It all starts with a well-maintained sample metadata file.**

## Maintain the sample metadata first

The sample metadata table is the authoritative record describing the samples
used in a project. It can be maintained in Excel, LibreOffice, Google Sheets,
R, a database, or another suitable tool. The software matters less than using a
consistent tabular structure and preserving the file carefully.

The tidy-data principles emphasized throughout
[R for Data Science](https://r4ds.hadley.nz) by Hadley Wickham,
Mine Çetinkaya-Rundel, and Garrett Grolemund are particularly useful:

* each row represents one sample;
* each column represents one variable;
* each cell contains one value;
* each type of observational unit is stored in its own table.

For example, sampling site, collection year, sex, sequencing batch, original
sample identifier, standardized sample identifier, and DArT `TARGET_ID` should
be separate columns. Do not combine several variables in one cell, use merged
cells or multiple header rows, or encode information only with colours, fonts,
or comments. Column names should be unique and stable, and missing values should
be represented consistently.

When using spreadsheet software, format identifier columns as text so that
leading zeros are preserved. Avoid formulas in the final exchange table and
export a tab-delimited text file for use with R. Keep the complete spreadsheet
or metadata table as the long-term record; exported strata files can always be
regenerated from it.

The metadata table is also the authoritative place to document original and
standardized identifiers, along with the reason or date for a name change. The
strata file should not be used to maintain that history.

## Prepare the strata file

Once the metadata are in good order, generate the strata file needed for the
genomic import or analysis. This small step prevents many difficult-to-diagnose
matching problems later. A strata file is a compact, tab-delimited table with
one row per sample and, for most formats, two required columns:

* `INDIVIDUALS` is the unique, stable name used for each sample;
* `STRATA` is the group to which the sample belongs for the current analysis.

For DArT data, a third column named `TARGET_ID` connects the names found in the
original DArT report to the standardized names in `INDIVIDUALS`.

### Standardize `INDIVIDUALS`

Good sample identifiers are meaningful, unique, and stable. Avoid identifiers
such as `1`, `2`, and `3`: they contain no context, are easily confused with row
numbers, and may change when a table is sorted. Prefer an identifier such as
`GWH-TAS-2025-001`, where each component has a documented meaning.

Useful practices include:

* record the identifier found in the genotype file and its standardized
  replacement in the sample metadata table;
* join existing metadata columns when necessary, for example species, sampling
  site, year, and a within-site sample number;
* use the same component order and separator for every sample;
* prefer hyphens (`-`) as separators; underscores can become visually unclear
  when identifiers are underlined or printed, and periods (`.`) can be altered
  by some R functions;
* avoid spaces, accented characters, slashes, and other punctuation;
* keep capitalization consistent and preserve meaningful leading zeros;
* verify that identifiers are unique, non-missing, and match the genotype data
  exactly.

Suppose the complete sample metadata are stored in `sample_metadata.tsv` with
columns named `SPECIES`, `SITE`, `YEAR`, `SAMPLE_NUMBER`, and `REGION`. A compact
strata file can be generated as follows:

```{r create-strata, eval=FALSE}
sample_metadata <- readr::read_tsv(
  "sample_metadata.tsv",
  col_types = readr::cols(.default = readr::col_character())
)

strata <- sample_metadata |>
  dplyr::mutate(
    INDIVIDUALS = paste(SPECIES, SITE, YEAR, REGION, SAMPLE_NUMBER, sep = "-"),
    STRATA = REGION
  ) |>
  dplyr::select(INDIVIDUALS, STRATA)

stopifnot(
  !anyNA(strata$INDIVIDUALS),
  !anyDuplicated(strata$INDIVIDUALS),
  !anyNA(strata$STRATA)
)

readr::write_tsv(strata, "strata.tsv")
```

Read identifier-like columns as character data so that values such as `001` do
not become `1`. Inspect the result before importing the genomic data:

```{r inspect-strata, eval=FALSE}
strata
dplyr::count(strata, STRATA)
```

### Renaming samples while reading a VCF

For a VCF, sample identifiers can be changed during import by adding `NEW_ID`
to the strata file. In this case, `INDIVIDUALS` must match the current sample
identifier in the VCF exactly, and `NEW_ID` supplies the standardized identifier
that will be written to the GDS object:

```text
INDIVIDUALS  NEW_ID           STRATA
Shark_1      GWH-TAS-2025-001 TAS
Shark_2      GWH-VIC-2025-001 VIC
Shark_3      GWH-TAS-2025-002 TAS
```

Generate this file from the authoritative metadata table rather than entering
the correspondence manually in the strata file:

```{r create-vcf-strata, eval=FALSE}
vcf_strata <- sample_metadata |>
  dplyr::transmute(
    INDIVIDUALS = VCF_ID,
    NEW_ID = STANDARD_ID,
    STRATA = REGION
  )

stopifnot(
  !anyNA(vcf_strata),
  !anyDuplicated(vcf_strata$INDIVIDUALS),
  !anyDuplicated(vcf_strata$NEW_ID)
)

readr::write_tsv(vcf_strata, "vcf_strata.tsv")
```

Here, `VCF_ID` and `STANDARD_ID` are columns maintained in
`sample_metadata.tsv`. If the identifiers already used in the VCF are suitable,
omit `NEW_ID` and use the regular two-column strata file.

### Keep `STRATA` simple

`STRATA` represents the grouping needed for the current analysis, not the
entire sample metadata record. Short acronyms such as `NTH` and `STH`, `POP1`
and `POP2`, or site codes such as `TAS` and `VIC` are preferable. Long labels
can make axes, legends, and facet labels unreadable when figure space is tight.
Use a single consistent vocabulary: do not mix values such as `STH`, `South`,
and `south`.

If a hierarchical label is needed, use a documented and consistent structure
such as `TAS-HOB` and `TAS-FRE`. Avoid packing every available variable into
`STRATA`. Keep collection date, sex, life stage, coordinates, sequencing batch,
and other descriptive fields in the complete sample metadata table. They can be
joined to analysis results later using `INDIVIDUALS` as the key.

Also check the following before saving the file:

* every row represents one sample;
* every sample needed for the analysis occurs exactly once;
* samples absent from the strata file are intentionally excluded;
* group names do not contain trailing spaces or inconsistent separators;
* the file is written as tab-delimited text rather than an Excel workbook.

A plain-text editor that can display invisible characters is useful for the
final inspection. For example, BBEdit can reveal tabs, spaces, and line endings,
making accidental spaces, mixed delimiters, and hidden characters easier to
find.

## Reading a VCF file

Suppose that `individuals.vcf.gz` contains the genotypes and `strata.tsv`
contains sample assignments. A minimal strata table has one row per individual:

```text
INDIVIDUALS  STRATA
GWH-TAS-2025-001  TAS
GWH-VIC-2025-001  VIC
GWH-TAS-2025-002  TAS
```

Use `read_vcf()` when VCF-specific control is wanted:

```{r read-vcf, eval=FALSE}
shark <- read_vcf(
  data = "individuals.vcf.gz",
  strata = "strata.tsv",
  filename = "shark",
  parallel.core = 8,
  verbose = TRUE
)
```

The function returns an open SeqArray GDS object. The GDS representation avoids
materializing a very large genotype table in memory and is the preferred format
for subsequent work in R.

### Preserve how variants were generated

`read_vcf()` stores a provenance table in the GDS metadata. It records the
original path, size, modification time, MD5 checksum, import time, detected
caller, exact VCF source value, imported INFO and FORMAT fields, and available
reference, contig, version, and command header lines. Keep the original VCF and
its workflow files as well: provenance stored in GDS supports interpretation,
but it is not a replacement for a maintained variant-calling pipeline.

This matters when a genome-scan peak, missingness pattern, or local PCA signal
is restricted to one caller, mapping strategy, sequencing batch, or reference
assembly. Such a pattern can be biological, technical, or both.

### Structural variants are not ordinary SNP dosages

When present, common annotations such as `SVTYPE`, `END`, `SVLEN`, `CIPOS`,
`CIEND`, `MATEID`, and `EVENT` are preserved during VCF import. This is the
first step toward fuller structural-variant support. Structural variants can
have uncertain breakpoints, symbolic ALT alleles, breakends, copy number, and
several forms of read support, so they should not all be forced into the SNP
`ALT_DOSAGE` representation.

Until a dedicated structural-variant representation is complete, keep these
annotations and the source VCF, and describe local signals cautiously. A local
PCA or LD haploblock may be consistent with a candidate inversion, but it may
also reflect a centromere, low recombination, assembly or mapping problems,
introgression, population-specific missingness, or another structural variant.
Use "candidate", "putative", or "inversion-associated haploblock" until
physical evidence supports a more specific structural interpretation.

VCF files that are not ready for parallel access may need to be sorted,
bgzip-compressed, and indexed. These preparation steps can require the optional
`bcftools` executable. Check its visibility with:

```{r check-bcftools, eval=FALSE}
Sys.which("bcftools")
genometranslator_dependencies()
```

Use `tidy_genome()` only when an in-memory tidy table is needed:

```{r vcf-to-tidy, eval=FALSE}
shark_tidy <- tidy_genome(
  data = shark,
  parallel.core = 8
)
```

## Reading DArT files

DArT reports have changed over time, and the package recognizes several related
layouts automatically:

* **1row**: one row per SNP marker, with genotypes encoded as `0`, `1`, `2`, or
  missing;
* **2rows**: two rows per marker representing presence or absence of the REF and
  ALT alleles;
* **counts**: two rows per marker containing read counts or depth for the REF and
  ALT alleles rather than genotype calls;
* **silico.dart**: dominant SilicoDArT presence/absence markers without REF and
  ALT alleles;
* **silico.dart.counts**: SilicoDArT presence/absence information accompanied by
  sequence counts.

For DArT input, prepare a three-column strata file. `TARGET_ID` must match the
identifier in the original DArT report, while `INDIVIDUALS` contains the stable,
standardized name that will be used after import:

```text
TARGET_ID  INDIVIDUALS       STRATA
1002451    GWH-TAS-2025-001  TAS
1002452    GWH-TAS-2025-002  TAS
1002453    GWH-VIC-2025-001  VIC
```

When the DArT identifiers are difficult to locate, extract them directly from
the report instead of editing the large genotype file manually:

```{r dart-target-id, eval=FALSE}
dart_ids <- extract_dart_target_id(
  data = "dart_counts.csv",
  write = FALSE
)
```

For DArT files, `INDIVIDUALS` must be generated by the user. This is the right
time to create meaningful, standardized identifiers. After verifying which
sample corresponds to each `TARGET_ID`, store that relationship permanently in
the complete sample metadata table. Then use the `TARGET_ID` values extracted
from the DArT report to select and join the corresponding metadata records. The
DArT strata file is a small export of the three fields needed for import:

```{r create-dart-strata, eval=FALSE}
unmatched_dart_ids <- dart_ids |>
  dplyr::anti_join(sample_metadata, by = "TARGET_ID")

stopifnot(nrow(unmatched_dart_ids) == 0L)

dart_strata <- dart_ids |>
  dplyr::left_join(sample_metadata, by = "TARGET_ID") |>
  dplyr::transmute(
    TARGET_ID,
    INDIVIDUALS = STANDARD_ID,
    STRATA = REGION
  )

stopifnot(
  !anyNA(dart_strata),
  !anyDuplicated(dart_strata$TARGET_ID),
  !anyDuplicated(dart_strata$INDIVIDUALS)
)

readr::write_tsv(dart_strata, "dart_strata.tsv")
```

Never infer this correspondence from row or column order alone. The optional
metadata inspection features of `extract_dart_target_id()` are described in
that function's documentation.

`detect_genomic_format()` identifies the file as DArT, while
`detect_dart_format()` reports the particular DArT layout:

```{r detect-dart, eval=FALSE}
detect_genomic_format("dart_counts.csv")
detect_dart_format("dart_counts.csv")
```

For a DArT count file, use:

```{r read-dart-counts, eval=FALSE}
dart_genome <- read_dart(
  data = "dart_counts.csv",
  strata = "dart_strata.tsv"
)
```

The specialized `read_dart()` documentation describes additional DArT columns,
metadata, and calibration options. Calling `read_dart()` directly is
recommended when those choices need to be controlled.

### Stable marker identifiers and coordinates

Imported marker metadata deliberately separates identity from genomic placement.
`MARKERS` is a stable identifier with three components separated by two
underscores. A de novo marker is written as
`DENOVO__<LOCUS>__<POS>`. For a reference marker, the exact sequence name is
kept in `CHROM`, while only the component embedded in `MARKERS` is made safe for
filenames and downstream code. For example, `CHROM = "NC_054479.1"` is encoded
as `NC_054479_1` inside `MARKERS`.

Consequently, genomic coordinates must always be taken from the `CHROM` and
`POS` columns, not reconstructed by splitting `MARKERS`. `LOCUS` retains the
original DArT/read locus and `COL` retains the SNP position within that source
sequence. Sanitizing two distinct chromosome names to the same marker component
is treated as an error rather than silently creating ambiguous identifiers.

## The generic read and write interface

When format-specific options are not required, `read_genome()` detects the
input and calls the corresponding reader with its defaults. Only the shared
`strata` argument needs to be supplied:

```{r read-genome, eval=FALSE}
genome <- read_genome(
  data = "individuals.vcf.gz",
  strata = "strata.tsv",
  parallel.core = 8,
  verbose = TRUE
)
```

The same entry point can be tried with a DArT, FSTAT, Genepop, GDS, PLINK, or
package-native genomic file, and with `genind`, `genlight`, or `gtypes` objects:

```{r read-genome-dart, eval=FALSE}
dart_genome <- read_genome(
  data = "dart_counts.csv",
  strata = "dart_strata.tsv"
)
```

`write_genome()` dispatches to a format-specific writer. Specify `output`
explicitly when the filename extension is ambiguous:

```{r write-genome, eval=FALSE}
write_genome(
  data = genome,
  output = "genepop",
  filename = "shark",
  verbose = TRUE
)

write_genome(
  data = genome,
  output = "vcf",
  filename = "shark_filtered",
  verbose = TRUE
)
```

The name `shark_filtered.vcf` should only be used if `genome` was filtered
before the call: the writer itself does not perform that filtering.

### Upgrade legacy `.gds.rad` files

New GDS files use the standard `.gds` extension regardless of whether their
source was DArT, VCF, PLINK, or another format. The source format and provenance
belong in the GDS metadata rather than in the filename.

Use `upgrade_genome_gds()` to migrate an older package or plain SeqArray GDS:

```{r upgrade-gds, eval=FALSE}
new_file <- upgrade_genome_gds("shark.gds.rad")
# Writes shark.gds and shark_upgrade_report.tsv
```

The original file is never modified. The upgrade copies the file atomically,
migrates `/radiator` metadata to `/genometranslator`, creates current-schema
nodes, reconstructs marker and individual metadata where possible, modernizes
legacy marker identifiers, inventories depth and likelihood nodes, and checks
metadata identifiers against the core SeqArray variants and samples. Previous
marker identifiers are retained in `ORIGINAL_MARKERS` when they change.

Information that was never stored cannot be recreated. The report therefore
distinguishes migrated, reconstructed, retained, unavailable, and failed or
warning checks. In particular, missing strata, source sequences, DP, AD, or
provenance must be supplied from the original files if they are required.

For a dry metadata-preserving conversion without identifier modernization, use:

```{r upgrade-gds-conservative, eval=FALSE}
new_file <- upgrade_genome_gds(
  "shark.gds.rad",
  modernize.markers = FALSE
)
```

When finished with an open GDS connection, close it with:

```{r close-genome, eval=FALSE}
write_genome(genome)
```

With no output requested, `write_genome()` retains its GDS-management behavior:
it synchronizes and closes the open connection.

## Translating in one step

`genome_translator()` combines `read_genome()` and `write_genome()`. For
example, translate a VCF directly to Genepop using the default behavior of both
format-specific functions:

```{r genome-translator, eval=FALSE}
genome_translator(
  data = "individuals.vcf.gz",
  strata = "strata.tsv",
  output = "genepop",
  filename = "shark",
  parallel.core = 8,
  verbose = TRUE
)
```

This compact interface is appropriate when the defaults are suitable. For more
control, split the operation into a specialized reader and writer:

```{r specialized-translation, eval=FALSE}
genome <- read_vcf(
  data = "individuals.vcf.gz",
  strata = "strata.tsv",
  vcf.stats = TRUE,
  parallel.core = 8
)

write_genepop(
  data = genome,
  filename = "shark"
)

write_genome(genome) # synchronize and close the GDS connection
```

Always consult the selected reader and writer documentation for format-specific
dependencies, validation rules, and data-preparation recommendations.
